RVD2: An ultra-sensitive variant detection model for low-depth targeted next-generation sequencing data

### 2014-01-09 doc/bioinformatics2014/document.tex
update manucript

\section{Availability:}
Text

\section{Contact:} \href{pjflaherty@wpi.edu}{pjflaherty}
\end{abstract}

@@

Increasingly, NGS is being used to interrogate mutations in heterogeneous clinical samples. For example, NGS-based non-invasive fetal DNA testing uses maternal blood sample to sequence the minority fraction of cell-free fetal DNA~\citep{Fan:2008di}. Infectious diseases such as HIV and influenza may contain many genetically heterogeneous sub-populations~\citep{Flaherty:2011ja, cushing2013rvd, Ghedin:2010ie}. DNA sequencing of individual regions of a solid tumor has revealed genetic heterogeneous within an individual sample~\citep{Navin:2010gu}.

@@

Recently, researchers have developed algorithms to call low-frequency or rare variants in heterogeneous samples.  VarScan2 combines algorithmic heuristics to call genotypes in the tumor and normal sample pileup data and then applies a Fisher's exact test on the read count data to detect a significant difference in the genotype calls~\citep{Koboldt:2012cg}. Strelka uses a hierarchical Bayesian approach to model the joint distribution of the allele frequency in the tumor and normal samples at each locus~\citep{Saunders:2012fh}. With the joint distribution available, one is able to identify locations with dissimilar allele frequencies. muTect uses a Bayesian posterior probability in its decision rule to evaluate the likelihood of a mutation~\citep{Cibulskis:2013ta}. RVD uses a hierarchical Bayesian model to capture the error structure of the data and call variants~\citep{}. However, that algorithm requires a very high read depth to estimate the sequencing error rate and call variants.

Several studies have compared the relative performance of these algorithms. \citet{spencer2013performance} demonstrated  that VarScan-somatic performed the best comparing with SAMtools, GATK and SPLINTER in detecting minor allele frequencies (MAFs) of 1\% to 8\%, with $ > $500× coverage required for optimal performance. However, \citet{spencer2013performance} also highlighted the fact that VarScan2 yielded more false positives at high read depth. \citet{Stead:2013fu} showed that VarScan-somatic outperformed Strelka had performance on-par with muTect in detecting a 5\% MAF for read depths between 100 and 1000. We test their performance on much lower MAF variants and across a wider range of coverage depths.

@@

\paragraph{Funding\textcolon}

>>>>

\section{Availability:}
\href{http://genomics.wpi.edu/rvd2/rvd2.html}{http://genomics.wpi.edu/rvd2/rvd2.html}

\section{Contact:} \href{pjflaherty@wpi.edu}{pjflaherty@wpi.edu}
\end{abstract}

@@

Increasingly, NGS is being used to interrogate mutations in heterogeneous clinical samples. For example, NGS-based non-invasive fetal DNA testing uses maternal blood sample to sequence the minority fraction of cell-free fetal DNA~\citep{Fan:2008di}. Infectious diseases such as HIV and influenza may contain many genetically heterogeneous sub-populations~\citep{Flaherty:2011ja, Ghedin:2010ie}. DNA sequencing of individual regions of a solid tumor has revealed genetic heterogeneous within an individual sample~\citep{Navin:2010gu}.

@@

Recently, researchers have developed algorithms to call low-frequency or rare variants in heterogeneous samples.  VarScan2 combines algorithmic heuristics to call genotypes in the tumor and normal sample pileup data and then applies a Fisher's exact test on the read count data to detect a significant difference in the genotype calls~\citep{Koboldt:2012cg}. Strelka uses a hierarchical Bayesian approach to model the joint distribution of the allele frequency in the tumor and normal samples at each locus~\citep{Saunders:2012fh}. With the joint distribution available, one is able to identify locations with dissimilar allele frequencies. muTect uses a Bayesian posterior probability in its decision rule to evaluate the likelihood of a mutation~\citep{Cibulskis:2013ta}. RVD uses a hierarchical Bayesian model to capture the error structure of the data and call variants~\citep{Flaherty:2011ja, cushing2013rvd}. However, that algorithm requires a very high read depth to estimate the sequencing error rate and call variants.

Several studies have compared the relative performance of these algorithms. \citet{spencer2013performance} demonstrated  that VarScan-somatic performed the best comparing with SAMtools, GATK and SPLINTER in detecting minor allele frequencies (MAFs) of 1\% to 8\%, with $ > $500× coverage required for optimal performance. However, \citet{spencer2013performance} also highlighted the fact that VarScan2 yielded more false positives at high read depth. \citet{Stead:2013fu} showed that VarScan-somatic outperformed Strelka had performance on-par with muTect in detecting a 5\% MAF for read depths between 100 and 1000. %We test their performance on much lower MAF variants and across a wider range of coverage depths.

@@

\paragraph{Funding\textcolon}  P.F. and Y.E. were supported by seed funding from Worcester Polytechnic Institute.
